From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-30 Pi+/paper/equivalence.tex
Working on sec 7

\section{Equivalence between \texorpdfstring{$\PiPlusLang$}{Pi} and \texorpdfstring{$\UFin$}{UFin}}~\label{sec:equivalence}

In this section, we use the semantics developed in the previous sections define a fragment $\PiPlusLang$ of $\PiLang$
and a normalised fragment of it $\PiHatLang$, interpret to $\UFin$ and back, and translate between the languages.

We only present the types and 1-combinators in the language, and the full set of 2-combinators is listed in the appendix
and the accompanying Agda code.

\vc{We can translate types to types, terms to terms, theorem: preserves types, preserves equations.}

\vc{Define maps quote/eval on 0,1,2-cells, and show commuting diagrams.}

\subsection{$\PiHatLang$}

We first present a language $\PiHatLang$ based on the theory of symmetric groups developed in~\cref{sec:finite}. The
types in the language are unary natural numbers, and combinators are allowed to perform adjacent transpositions.

@@

\subsection{$\PiPlusLang$}

Now, we present the additive fragment of $\PiLang$, called $\PiPlusLang$.

\begin{proposition}
  We can form a weak 2-category $\PiPlusCat$ with
  \begin{itemize}
    \item $\PiPlusLang$ types for 0-cells,
    \item for $X, Y : \UPlus$, a collection of 1-cells $X \iso Y$,
    \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Iso q$.
  \end{itemize}
\end{proposition}

\begin{proposition}
  There is a symmetric monoidal structure on $\PiPlusCat$, with $\zerot$ for the unit, and $+$ for the tensor.
\end{proposition}

\subsection{$\PiLang$}

Finally, We show how to translate $\PiLang$ programs to $\PiHatLang$ programs.

\begin{definition}
  \begin{align*}
    \evalt_{0} & : U \to \Nat                                               \\
    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
  \end{align*}
\end{definition}

To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$ and quote it to $\PiPlusLang$.

>>>>

\section{Going from \texorpdfstring{$\PiLang$}{Pi} to \texorpdfstring{$\UFin$}{UFin} and back}~\label{sec:equivalence}

% https://q.uiver.app/?q=WzAsNCxbMCwwLCJcXFBpTGFuZyJdLFsyLDAsIlxcUGlQbHVzTGFuZyJdLFs0LDAsIlxcUGlIYXRMYW5nIl0sWzYsMCwiXFxVRmluIl0sWzAsMSwiXFxldmFsdCJdLFsxLDIsIlxcZXZhbGgiLDAseyJjdXJ2ZSI6LTR9XSxbMiwzLCJcXGV2YWx0IiwwLHsiY3VydmUiOi00fV0sWzIsMSwiXFxxdW90ZWgiLDAseyJjdXJ2ZSI6LTR9XSxbMywyLCJcXHF1b3RldCIsMCx7ImN1cnZlIjotNH1dXQ==
\[\begin{tikzcd}
    \PiLang && \PiPlusLang && \PiHatLang && \UFin
    \arrow["\evalt", from=1-1, to=1-3]
    \arrow["\evalp", curve={height=-24pt}, from=1-3, to=1-5]
    \arrow["\evalh", curve={height=-24pt}, from=1-5, to=1-7]
    \arrow["\quotep", curve={height=-24pt}, from=1-5, to=1-3]
    \arrow["\quoteh", curve={height=-24pt}, from=1-7, to=1-5]
  \end{tikzcd}\]

In this section, we consider the additive fragment $\PiPlusLang$ of $\PiLang$. This is the language that we interpret to
$\UFin$ and back, using the tools developed in the previous sections. Further, we go through an intermediate step of the
language $\PiHatLang$, which is a simplified variant of $\PiPlusLang$ that uses adjacent transpositions for combinators,
but still preserving all the required structure.

We present the types and 1-combinators of $\PiPlusLang$ and $\PiHatLang$ in~\cref*{fig:piplus,fig:pihat} respectively,
eliding the 2-combinators for brevity. \note{We enforce that there is a unique 2-combinator between compatible
  1-combinators, by introducing a level 3, which is trivial in the same way as in the case of $\PiLang$, as described
  before.} They can be found in the appendix and in the accompanying Agda code.

The translations between the languages are defined separately on types, 1-combinators, and 2-combinators. Following the
terminology of Normalisation by Evaluation, the translations from the left to the right, going from the syntax towards
the semantics, are called $\evalt$ and the translations the other way are called $\quotet$.

To state our results formally, we organise the syntax for each language using a technical device, called a syntactic
category. Since the language have higher-dimensional structure, the syntactic categories here are actually
(locally-strict) weak 2-categories. We define them formally in the appendix, and only state our results here.

For each of the $\PiLang$, $\PiPlusLang$ and $\PiHatLang$ languages, their syntactic categories, respectively $\PiCat$,
$\PiPlusCat$ and $\PiHatCat$, have 0-cells for types, 1-cells for 1-combinators, and 2-cells for 2-combinators. We can
show that these are $(2,0)$-categories, since all the 1-cells and 2-cells are invertible. They are also locally-strict,
or locally-posetal, because there is at most one 2-cell between compatible 1-cells.

\begin{toappendix}
  \todo{What is U?}
  \begin{proposition}
    We can form  weak 2-category $\PiCat$ with
    \begin{itemize}
      \item $\PiLang$ types for 0-cells,
      \item for $X, Y : U$, a collection of 1-cells $X \iso Y$,
      \item for $p, q : X \iso Y$, a collection of 2-cells $p \Iso q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form  weak 2-category $\PiPlusCat$ with
    \begin{itemize}
      \item $\PiPlusLang$ types for 0-cells,
      \item for $X, Y : \UPlus$, a collection of 1-cells $X \isop Y$,
      \item for $p, q : X \isop Y$, a collection of 2-cells $p \Isop q$.
    \end{itemize}
  \end{proposition}

  \begin{proposition}
    We can form  weak 2-category $\PiHatCat$ with
    \begin{itemize}
      \item $\PiHatCat$ types for 0-cells,
      \item for $X, Y : \UHat$, a collection of 1-cells $X \isoh Y$,
      \item for $p, q : X \isoh Y$, a collection of 2-cells $p \Isoh q$.
    \end{itemize}
  \end{proposition}
\end{toappendix}

We use the $\evalt/\quotet$ translation maps to construct functors between these categories. We only name the maps on
the 0, 1, and 2-cells -- the coherences hold by definition or by calculation, which is shown in our accompanying Agda
code. We use these functors to state our results establishing the equivalences between the languages.

\subsection{$\PiLang$ to $\PiPlusLang$}

First, we show how to translate $\PiLang$ programs to $\PiPlusLang$, which is the additive fragment of $\PiLang$. The
syntax is given in~\cref{fig:piplus}.

$\PiLang$ has two 0-ary type constructors, and two binary type constructors -- the additive tensor product and the
multiplicative one. $\PiPlusLang$ has all the type constructors of $\PiLang$ except multiplication. However, we will
show how to recover the multiplicative structure, by defining multiplication as repeated addition. We encode $\times$ in
terms of $+$ as follows.

\begin{definition}[$\times : \UPlus \to \UPlus \to \UPlus$]
  \begin{align*}
    \zerot \times Y      & \defeq \zerot                      \\
    \onet \times Y       & \defeq Y                           \\
    (X_1 + X_2) \times Y & \defeq X_1 \times Y + X_2 \times Y
  \end{align*}
\end{definition}

\begin{proposition}
  There are two symmetric monoidal structures on $\PiPlusCat$, given by $(\zerot, +)$ and $(\onet, \times)$, with
  $\times$ distributing over $+$, giving it a rig structure, as described in \todo{cite earlier section}.
\end{proposition}
\todo{either prove it or cite something}

Using this rig structure, we translate $\PiLang$ to $\PiPlusLang$, constructing a rig functor from $\PiCat$ to
$\PiPlusCat$.

\begin{definition}
  \begin{align*}
    \evalt_{0} & : U \to \UPlus                                             \\
    \evalt_{1} & : (c : X \iso Y) \to \evalt_{0}(X) \iso \evalt_{0}(Y)      \\
    \evalt_{2} & : (\alpha : p \Iso q) \to \evalt_{1}(p) \Iso \evalt_{1}(q) \\
  \end{align*}
\end{definition}

\subsection{$\PiPlusLang$ to $\PiHatLang$}

Next, we show how to translate $\PiPlusLang$ programs to $\PiHatLang$ and back. $\PiHatLang$ is a simplified variant of
$\PiPlusLang$, with (unary) natural numbers for types, constructed by $\zero$ and $\suc$, 1-combinators generated by
adjacent transpositions, and an appropriate set of 2-combinators. The syntax, again omitting 2-combinators, is given
in~\cref{fig:pihat}. \vc{Something about PROPs.}

\subsection{$\PiHatLang$ to $\UFin$}

\vc{Define maps quote/eval on 0,,2-cells, and show commuting diagrams.}

@@

\subsection{$\PiLang$ circuit normalization}

To normalise a $\PiLang$ circuit, we translate it to $\PiHatLang$, compute a permutation, and quote back it to
$\PiPlusLang$.
